F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-02-17 icfp-2015/paper.tex
Writing something on semantics

@@

@@

\item We introduce the notion of relativized type safety and show how it holds for the code
  written using our type providers (Section~\ref{sec:safety}). Although we focus on our specific
  case, this demonstrates a new way of thinking about ML-style type safety that is needed
  in the context of web.
\end{itemize}

@@

More details about the full type erasure mechanism can be found in the paper on F\# type 
providers [?]. We present a simplified model when discussing safety in Section~\ref{sec:safety}.

@@

Note that the subtyping relation between structural types $\sigma$ does not map to a subtyping 
relation between F\#  types $\tau$. However, it does hold at runtime, meaning that an operation on 
a supertype can be performed on its subtypes. We make the type provider mechanism and runtime aspects 
precise in Section~\ref{sec:safety} when discussing safety.

@@

\begin{figure*}[t]

@@

% all rules are symmetric and reflexive
\begin{equation*}
\hspace{4.75em}
\inference[(sym)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
%
\qquad\quad
\begin{array}{l}
 \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.75em]
 % top type
 \textnormal{\footnotesize{(top)}}\;\; \top \tsep \sigma \vdash \sigma
\end{array}
%
\qquad\quad 
% null and nullable
 \begin{array}{l}
 \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.75em]
 \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
 \end{array}
\end{equation*}

\caption{Inference judgements that define the common supertype relation}
\label{fig:subtyping-cst}
\end{figure*}

@@

\section{Relativized type safety}
\label{sec:safety}
~

% - Featherweight F\#
% - Generating FF\# from inferred types
% - Evaluation reduction
% - Safety property

@@

@@

>>>>

% TODO: The model here does not actually use type erasure!!

@@

\newcommand{\reduce}{\rightarrow}

@@

\item We provide a minimal formal model (Section~\ref{sec:formal}) and use it to prove a
  relativized notion of type safety for our type providers (Section~\ref{sec:safety}). Although 
  we focus on our specific case, the relativized safety property demonstrates a new way of 
  thinking about ML-style type safety that is needed in the context of web.
\end{itemize}

@@

The type erasure mechanism is explained in our simplified formal model in Section~\ref{sec:formal}.
More details about the full implementation in F\# can be found in the paper on F\# type 
providers [?].

@@

Note that the subtyping relation between structural types $\sigma$ does not map to a subtyping 
relation between F\#  types $\tau$. However, it does hold at runtime, meaning that an operation on 
a supertype can be performed on its subtypes. We make the type provider mechanism and runtime aspects 
precise in Section~\ref{sec:formal} before discussing safety in Section~\ref{sec:safety}.

@@

\begin{figure*}[t]
{\small{

@@

% all rules are symmetric and reflexive
\begin{equation*}
\hspace{4.75em}
\inference[(sym)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
%
\qquad\quad
\begin{array}{l}
 \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.75em]
 % top type
 \textnormal{\footnotesize{(top)}}\;\; \top \tsep \sigma \vdash \sigma
\end{array}
%
\qquad\quad 
% null and nullable
 \begin{array}{l}
 \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.75em]
 \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
 \end{array}
\end{equation*}
}}
\caption{\small{Inference judgements that define the common supertype relation}}
\label{fig:subtyping-cst}
\end{figure*}

@@

\begin{figure*}
{\small{
\begin{equation*}
\hspace{3em}
\begin{array}{l}
 \ident{asBool}(i) \reduce \kvd{true}  \quad(i = 1)\\
 \ident{asBool}(d) \reduce \kvd{true}  \quad(d = 1.0)\\
 \ident{asBool}(f) \reduce \kvd{true}  \quad(f = 1.0)\\
\end{array}
\qquad
\begin{array}{l}
 \ident{asBool}(i) \reduce \kvd{false} \quad(i = 0)\\
 \ident{asBool}(d) \reduce \kvd{false} \quad(d = 0.0)\\
 \ident{asBool}(f) \reduce \kvd{false} \quad(f = 0.0)\\
\end{array}
\qquad
\begin{array}{l}
 \ident{asBool}(s) \reduce \kvd{true}  \quad(s=\str{true} \vee s=\str{1}) \\
 \ident{asBool}(s) \reduce \kvd{false} \quad(s=\str{false} \vee s=\str{0}) \\
 \ident{asBool}(b) \reduce b\quad(\textnormal{when}~b=\kvd{true}\vee b=\kvd{false})
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{-1em}
\begin{array}{l}
 \ident{asInt}(i) \,\reduce i\\
 \ident{asInt}(d) \reduce i\quad(i = \lfloor d \rfloor)\\
 \ident{asInt}(f) \reduce i\quad(i = \lfloor f \rfloor) \\
 \ident{asInt}(s) \,\reduce i\quad(s~\text{represents}~i) \\
 \ident{asInt}(\kvd{true}) \reduce 1\\
 \ident{asInt}(\kvd{false}) \reduce 0\\
\end{array}
\quad
\begin{array}{l}
 \ident{asDec}(i) \,\reduce d\quad(d = i)\\
 \ident{asDec}(d) \reduce d\\
 \ident{asDec}(f) \reduce d\quad(d = \lfloor f \rfloor) \\
 \ident{asDec}(s) \,\reduce d\quad(s~\text{represents}~d) \\
 \ident{asDec}(\kvd{true}) \reduce 1.0\\
 \ident{asDec}(\kvd{false}) \reduce 0.0\\
\end{array}
\quad
\begin{array}{l}
 \ident{asFloat}(i) \,\reduce f\quad(f = i)\\
 \ident{asFloat}(d) \reduce f\quad(f = d)\\
 \ident{asFloat}(f) \reduce f\\
 \ident{asFloat}(s) \,\reduce f\quad(s~\text{represents}~f) \\
 \ident{asFloat}(\kvd{true}) \reduce 1.0\\
 \ident{asFloat}(\kvd{false}) \reduce 0.0\\
\end{array}
\quad
\begin{array}{l}
 \ident{asStr}(i) \,\reduce s\quad(s~\text{represents}~i)\\
 \ident{asStr}(d) \reduce s\quad(s~\text{represents}~d)\\
 \ident{asStr}(f) \reduce s\quad(s~\text{represents}~f)\\
 \ident{asStr}(s) \,\reduce s\\
 \ident{asStr}(\kvd{true}) \reduce \str{true}\\
 \ident{asStr}(\kvd{false}) \reduce \str{false}\\
\end{array}
\end{equation*}
%
\begin{equation*}
\hspace{8em}
\begin{array}{l}
 \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]\\
 \ident{getChildren}(\kvd{null}) \reduce [~]\\
\end{array}
\qquad\qquad\qquad
\begin{array}{l}
 \ident{getField}(\nu,\nu_i,\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
 \ident{getField}(\epsilon,\nu_i,\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
\end{array}
\end{equation*}
}}
\caption{\small{Reduction rules for conversion functions}}
\label{fig:op-conversions}
\end{figure*}

% TODO: We'll need more things here for working with options 
% Also, 'getField' should return 'null' if the field does not exist
% (which we will presumably handle nicely thanks to handling of options)

\section{Formalising type providers}
\label{sec:formal}

In this section, we build the necessary theoretical framework for proving the relativised type safety 
property of F\# Data type providers in Section~\ref{sec:safety}. We start by discussing the runtime 
representation of structured values and runtime conversions (Section~\ref{sec:formal-convert}). Then 
we embed these into a formal model of an F\# subset that is necessary for discussing type providers 
(Section~\ref{sec:formal-ff}) and we describe how F\# Data type providers turn an inferred 
structural type into actual F\# code (Section~\ref{sec:formal-tp}).

\subsection{Structural values and conversions}
\label{sec:formal-convert}

In the model presented here, we represent JSON, XML and CSV documents using the same 
\emph{structural value}\footnote{Here, we diverge from the actual implementation in F\# Data which 
uses a different implementation for each format (reusing existing libraries).}. 
A structural value are first-order and can be one of the primitive values 
(Boolean, string, integer, decimal and float), a missing value (\kvd{null}), a record 
(named or unnamed) and a collection:
%
\begin{equation*}
\begin{array}{lcl}
 s  &\narrow{=}& i \lsep d \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
    &\narrow{|}& [s_1; \ldots; s_n] \\[0.1em]
    &\narrow{|}& \{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}\\[0.1em]
    &\narrow{|}& \nu~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \} \\
\end{array}
\end{equation*}

\noindent
As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
value \num{1} can be treated as \kvd{true}). To capture these, the operational semantics accesses 
values using the following operations:
%
\begin{equation*}
\begin{array}{lcl}
 op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s) \\
    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)
\end{array}
\end{equation*}
%
The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
to note that the relativised property we intend to prove about F\# Data provides \emph{sufficient}, but
not \emph{necessary} conditions. This means that a type provided based on sample is guaranteed to work 
correctly on certain inputs, but may also handle several additional inputs.

The conversion functions follow this approach. They perform conversions required by the subtyping relation
(\emph{e.g.}~integer is converted to decimal or a float) but also attempt additional conversions.

% TODO: Finish this

\subsection{Featherweight F\#}
\label{sec:formal-ff}

Types are primitive types or a class or a StructVal, which represents different structured values

L
\begin{equation*}
\begin{array}{l}
 \tau = \ident{int}\lsep\ident{decimal}\lsep\ident{float}\lsep\ident{bool}\lsep\ident{string}\lsep \ident{StructVal} \lsep C \\[0.2em]
 L = \kvd{type}~C(\overline{f:\tau}) = \overline{M} \\[0.2em]
 M = \kvd{member}~N:\tau=e \\[0.2em]
 v = ...
\end{array}
\end{equation*}

.. todo -- evaluation goes here too

\subsection{Type providers}
\label{sec:formal-tp}

How we turn inferred structural type into F\# type

\section{Relativized type safety}
\label{sec:safety}

% - Safety property

@@

Error handling - solve issues just add another example

@@

  - Prime example of type providers, explores relativized type safety
  - Used by industry, downloads, tested, etc. 
  - Predictable yet powerful type inference
